## Takeaways from Chapter 5: Gaussian Process Regression

Based on [Chapter 5 of Surrogates](https://bookdown.org/rbg/surrogates/chap5.html):

1.  **Gaussian Process (GP) Definition**: A GP is a collection of random variables, any finite number of which have a joint multivariate normal distribution. It is determined by its mean function $\mu(x)$ and covariance function $\Sigma(x, x')$.
2.  **Zero-Mean Assumption**: Location invariant zero-mean GP modeling is the default in computer surrogate modeling. Any mean trend is often subtracted or modeled separately.
3.  **Covariance Kernels**: The "action" is in the covariance. A common choice is the inverse exponentiated squared Euclidean distance (Gaussian kernel):
    $$ \Sigma(x, x') = \tau^2 \exp \left\{ - \frac{||x - x'||^2}{\theta} \right\} $$
    where $\tau^2$ is the scale/amplitude and $\theta$ is the lengthscale (controlling "wiggliness").
4.  **Nugget Effect**: To handle noise or numerical stability, a "nugget" parameter $g$ is added to the diagonal of the covariance matrix:
    $$ \Sigma_n = \tau^2 (C_n + gI_n) $$
    This allows the model to separate signal from noise (smoothing instead of interpolation).
5.  **Prediction (Kriging)**: Predictions at new locations $X$ given training data $X_n, Y_n$ follow conditional multivariate normal equations. The predictive mean is a linear predictor of the observations (BLUP).
6.  **Hyperparameter Estimation**: Parameters like $\tau^2, \theta, g$ are typically estimated via Maximum Likelihood Estimation (MLE) or Bayesian methods.

## Example: Noisy 1d Sinusoid (with Nugget estimation)

This example extends the simple 1d GP to include noise and estimates the nugget parameter $g$.

```{r noisy_1d}
library(plgp)
library(mvtnorm)

# Data generation with noise
n_noisy <- 16 # Doubles the previous example size for better noise handling
X_noisy <- matrix(seq(0, 2*pi, length=n_noisy/2), ncol=1)
X_noisy <- rbind(X_noisy, X_noisy) # Replicates
y_noisy <- 5*sin(X_noisy) + rnorm(n_noisy, sd=1)

D_noisy <- distance(X_noisy) # pairwise squared Euclidean distances

# Negative Log-Likelihood function for Nugget (g)
nlg <- function(g, D, Y) {
  n <- length(Y)
  K <- exp(-D) + diag(g, n)
  Ki <- solve(K)
  ldetK <- determinant(K, logarithm=TRUE)$modulus
  ll <- - (n/2)*log(t(Y) %*% Ki %*% Y) - (1/2)*ldetK
  return(-ll)
}

# Optimize for g
g_hat <- optimize(nlg, interval=c(sqrt(.Machine$double.eps), var(y_noisy)), D=D_noisy, Y=y_noisy)$minimum
print(paste("Estimated nugget g:", g_hat))

# Re-calculate quantities with estimated g
K <- exp(-D_noisy) + diag(g_hat, n_noisy)
Ki <- solve(K)
tau2hat <- drop(t(y_noisy) %*% Ki %*% y_noisy / n_noisy)

# Prediction
XX <- matrix(seq(-0.5, 2*pi + 0.5, length=100), ncol=1)
DXX <- distance(XX)
DX <- distance(XX, X_noisy)

KX <- exp(-DX)
KXX <- exp(-DXX) + diag(g_hat, nrow(DXX)) # Nugget on diagonal for predictive var

mup <- KX %*% Ki %*% y_noisy
Sigmap <- tau2hat * (KXX - KX %*% Ki %*% t(KX)) # Predictive Covariance

# Quantiles
q1 <- mup + qnorm(0.05, 0, sqrt(diag(Sigmap)))
q2 <- mup + qnorm(0.95, 0, sqrt(diag(Sigmap)))

# Plotting
# "Sigma.int" subtracts nugget for "clean" signal visualization (optional 'trick')
Sigma.int <- tau2hat * (exp(-DXX) + diag(sqrt(.Machine$double.eps), nrow(DXX)) - KX %*% Ki %*% t(KX))
YY_plot <- rmvnorm(100, mup, Sigma.int)

matplot(XX, t(YY_plot), type="l", lty=1, col="gray", xlab="x", ylab="y", main="Noisy GP Fit")
points(X_noisy, y_noisy, pch=20, cex=1)
lines(XX, mup, lwd=2)
lines(XX, 5*sin(XX), col="blue") # True function
lines(XX, q1, lwd=2, lty=2, col=2)
lines(XX, q2, lwd=2, lty=2, col=2)
```

## Using Libraries: `laGP`

Instead of "cutting-and-pasting" matrix algebra, libraries like `laGP` provide optimized implementations (often in C).

```{r laGP_example}
library(laGP)

# Setup toy data (2D Friedman function example subset)
# Using the noisy 1d data from above for simplicity in this demo
X_laGP <- X_noisy
y_laGP <- y_noisy

# Initialization
# newGPsep for separable (anisotropic) GP, newGP for isotropic
# Here we use 1D so they are similar, but laGP separates lengthscale d and nugget g
gpi <- newGP(X_laGP, y_laGP, d=0.1, g=0.1*var(y_laGP), dK=TRUE)

# MLE estimation
# mleGP searches for optimal parameters
mle <- mleGP(gpi, tmin=sqrt(.Machine$double.eps), tmax=10)
print(mle$theta) # Estimated d


# Prediction
p <- predGP(gpi, XX, lite=TRUE) # lite=TRUE returns mean and var (s2)

# Cleanup
deleteGP(gpi)

# Visualizing laGP result
plot(X_laGP, y_laGP, pch=20, main="laGP Prediction")
lines(XX, p$mean, lwd=2, col="purple")
lines(XX, p$mean + 1.96*sqrt(p$s2), lty=2, col="purple")
lines(XX, p$mean - 1.96*sqrt(p$s2), lty=2, col="purple")
```

# Detailed Reference Section (Memo)

## 1. Mathematical Formulation

### Covariance Kernels
The Gaussian (Squared Exponential) kernel is the standard choice:

**Isotropic (Scalar $\theta$):**
$$ K(x, x') = \exp \left\{ - \frac{||x - x'||^2}{\theta} \right\} + g\delta_{x,x'} $$

**Separable/Anisotropic (Vector $\theta$):**
Crucial for high-dimensional inputs where sensitivity varies by dimension.
$$ K(x, x') = \exp \left\{ - \sum_{k=1}^m \frac{(x_k - x'_k)^2}{\theta_k} \right\} + g\delta_{x,x'} $$

### Negative Log-Likelihood (Concentrated)
For maximizing hyperparameters given data $D_n = \{X_n, Y_n\}$:
$$ \ell(g, \theta) = c - \frac{n}{2} \log(Y_n^\top K_n^{-1} Y_n) - \frac{1}{2} \log |K_n| $$
*Note: This concentrates out the scale parameter $\tau^2$, allowing optimization of only $\theta$ and $g$.*

## 2. Algorithm 5.1: GP Regression Steps

1.  **Define Kernel**: Choose correlation structure $K(\cdot, \cdot)$ (e.g., Gaussian, Matérn) with hyperparameters $\theta$ (lengthscale) and $g$ (nugget).
2.  **Likelihood**: Formulate the concentrated log-likelihood $\ell(\theta, g)$.
3.  **Optimize**: Use a numerical optimizer (e.g., `optim` with `L-BFGS-B`) to find Maximum Likelihood Estimates (MLE) $\hat{\theta}, \hat{g}$.
    *   *Tip*: Using gradients $\nabla \ell$ significantly speeds up convergence.
4.  **Predict**:
    *   Compute covariance matrices utilizing $\hat{\theta}, \hat{g}$:
        *   $K$: Training-Training covariance (with nugget).
        *   $K(X, \cdot)$: Testing-Training covariance (no nugget).
        *   $K(X, X)$: Testing-Testing covariance (with nugget for predictive uncertainty).
    *   Solve linear system $K^{-1}Y$ (often using Cholesky decomposition for stability).
    *   Apply conditional mean/variance equations:
        $$ \mu(X) = K(X, X_n) K_n^{-1} Y_n $$
        $$ \Sigma(X) = \tau^2 [ K(X, X) - K(X, X_n) K_n^{-1} K(X_n, X) ] $$

## 3. High-Dimensional Modeling with `laGP`

Standard GPs scale poorly ($O(N^3)$). The `laGP` library addresses this via:
*   **C Implementation**: Optimized linear algebra.
*   **Separable Kernels**: `newGPsep` allows optimizing unique $\theta_k$ for each input dimension.
*   **Local Approximation (Chapter 9)**: For massive datasets, `laGP` can fit local GPs to subsets of data (neighborhoods) rather than the fill dataset, enabling scaling.

**Code Pattern for Separable GP in `laGP`:**
```r
library(laGP)
# 1. Initialize (C-side object allocation)
gpi <- newGPsep(X, y, d=0.1, g=0.1*var(y), dK=TRUE)

# 2. MLE Optimization
mle <- mleGPsep(gpi, param="both", tmin=c(eps, eps), tmax=c(10, var(y)))

# 3. Predict
p <- predGPsep(gpi, XX)

# 4. Cleanup (Crucial!)
deleteGPsep(gpi)
```

## 4. Limits of Stationarity (Motivation for Deep GPs)
Standard GPs assume **Stationarity**: the covariance depends only on distance $x - x'$.
*   **Problem**: If a function is wiggly in one region and flat in another, a stationary GP struggles. It will either:
    *   Over-smooth the wiggly part (large $\theta$).
    *   Over-fit noise in the flat part (small $\theta$).
*   **Solution**: Non-stationary models.
    *   *Treed GPs*: Partition space and fit separate GPs.
    *   *Deep GPs*: Composition of GPs to learn complex mappings (warping the input space).

## 5. Comparison: GP vs Others
*   **GP vs Splines**: GPs generalize better in higher dimensions ($>2-3D$) where tensor-product splines explode in complexity.
*   **GP vs MARS**: GPs typically provide lower RMSE but are computationally heavier ($O(N^3)$ vs $O(N)$ or $O(N^2)$). GPs provide "honest" uncertainty quantification (UQ).

## 6. Definitions for Future Methods
*   **Isotropic**: Correlation decays at same rate in all directions.
*   **Anisotropic/Separable**: Correlation decay rates ($\theta_k$) differ per dimension.
*   **Nugget ($g$)**: Represents measurement error or micro-scale variability. Effectively regularizes the matrix inversion.



